---
title: "Holmusk_Challenge"
output: html_document
date: "2023-06-02"
---
### Load data ###
```{r}
bill_amt = read.csv("bill_amount.csv")
bill_id = read.csv("bill_id.csv")
clinical = read.csv("clinical_data.csv")
demo = read.csv("demographics.csv")
```


### Clean data ### 
```{r}
library(dplyr)
```

1. Data frame: bill_amt
```{r}
str(bill_amt)
colSums(is.na(bill_amt)) 
```

2. Data frame: bill_id 
```{r}
str(bill_id)
bill_id$date_of_admission = as.Date(bill_id$date_of_admission, format = "%Y-%m-%d") 
colSums(is.na(bill_id)) 
```

3. Data frame: clinical
```{r}
str(clinical)

colSums(is.na(clinical))  

# date_of_admission, date_of_discharge 
clinical$date_of_admission = as.Date(clinical$date_of_admission, format = "%Y-%m-%d") 
clinical$date_of_discharge = as.Date(clinical$date_of_discharge, format = "%Y-%m-%d")
```

```{r}
# medical_history_1
table(clinical$medical_history_1)
clinical$medical_history_1 = as.factor(clinical$medical_history_1)
summary(clinical$medical_history_1)

# medical_history_2 - 233 NA  
table(clinical$medical_history_2)
clinical$medical_history_2 = as.factor(clinical$medical_history_2)
str(clinical$medical_history_2)
 
# medical_history_3
table(clinical$medical_history_3)
clinical["medical_history_3"][clinical["medical_history_3"] == "No"] = "0"
clinical["medical_history_3"][clinical["medical_history_3"] == "Yes"] = "1"
clinical$medical_history_3 = as.factor(clinical$medical_history_3)
str(clinical$medical_history_3)

# medical_history_4, 5, 6, 7
# medical_history_5 - 304 NA  
clinical[7:10] = lapply(clinical[7:10], as.factor)
str(clinical)
```

```{r}
# preop_medication_1 to 6 
clinical[11:16] = lapply(clinical[11:16], as.factor)
```

```{r}
# symptom_1 to 5
clinical[17:21] = lapply(clinical[17:21], as.factor)
```

```{r}
# rename patient id col in clinical 
clinical = rename(clinical, patient_id = "id") 
```

4. Data frame: demo
```{r}
str(demo)

table(demo$gender) 
demo["gender"][demo["gender"] == "f"] = "Female"
demo["gender"][demo["gender"] == "m"] = "Male"

table(demo$race)
demo["race"][demo["race"] == "chinese"] = "Chinese"
demo["race"][demo["race"] == "India"] = "Indian"  

table(demo$resident_status)
demo["resident_status"][demo["resident_status"] == "Singapore citizen"] = "Singaporean"

demo[2:4] = lapply(demo[2:4], as.factor)

demo$date_of_birth = as.Date(demo$date_of_birth, format = "%Y-%m-%d")   

colSums(is.na(demo))
``` 


### Handle repeats & duplicates ###
```{r}
# check for duplicated rows  
sum(duplicated(bill_amt))
sum(duplicated(bill_id))
sum(duplicated(clinical))
sum(duplicated(demo))   
```  
There are no duplicated rows. 

```{r}
bill_combined = inner_join(bill_id, bill_amt, by = "bill_id")  
```

Check for split payments on the same occasion  
```{r}   
df1 = bill_combined %>%
  add_count(patient_id, date_of_admission) %>%
  filter(n>1) %>%
  distinct() 
```
Each payment is split into 4 bills  

  
Combine 4 bills into one for each patient on each occasion
```{r}
df2 = bill_combined %>% 
  group_by(patient_id, date_of_admission) %>%
  summarise_at(vars(amount), funs(sum)) %>% 
  ungroup()

bill_final = df2 %>% rename(total_bill = "amount")
```

Next, find out how many patients have repeated admissions & find number of unique patients.  
```{r}
bill_final %>% distinct(patient_id)
clinical %>% distinct(patient_id) 
demo %>% distinct(patient_id)
```
Number of unique patients is 3,000

```{r}
num_admissions = bill_final %>%
  add_count(patient_id)
```

```{r}
num_admissions = num_admissions %>% rename(total_num_admissions = n)
```

```{r}
num_admissions %>% filter(total_num_admissions==4)
num_admissions %>% filter(total_num_admissions==3)
num_admissions %>% filter(total_num_admissions==2)
num_admissions %>% filter(total_num_admissions==1)
```
4 admits: 1 patient
3 admits: 57/3 = 19 patients
2 admits: 718/2 = 359 patients
1 admit: 2,621 patients  


### Join data frames ###  
```{r}
final_df = full_join(num_admissions, clinical, by = c("patient_id", "date_of_admission"))  
final_df = left_join(final_df, demo, by = "patient_id") 
```    


### Feature engineering ###
```{r}
# length of stay
final_df = mutate(final_df, len_stay = date_of_discharge - date_of_admission)
final_df$len_stay = as.numeric(final_df$len_stay) 

# BMI
final_df = mutate(final_df, BMI = weight/((height/100)^2))  

# age upon admission
final_df = mutate(final_df, admission_age = date_of_admission - date_of_birth) 
final_df$admission_age = as.numeric(final_df$admission_age)
final_df = mutate(final_df, admission_age = admission_age/365)
```

```{r}
# nth admission 
final_df = arrange(final_df, desc(total_num_admissions), patient_id, date_of_admission) 
filter(final_df, total_num_admissions != 1) #779 rows

final_df[, "nth_admission"] = 1

for (i in 2:779) {  
  if (final_df[i,1] == final_df[i-1,1]) { # same patient
  final_df[i,36] = final_df[i-1, 36] + 1   
  }
}  

View(final_df %>% select(patient_id, date_of_admission, nth_admission, total_num_admissions)) 
```


### Handle NA values ### 
Show rows with NA values 
```{r}
NA_vals = final_df[!complete.cases(final_df), ]
summary(NA_vals)
```
502 rows  
 
```{r}
summary(final_df) 
```

Based on a visual analysis of the summary statistics of the full dataset & the subset containing only NA values, they share very similar means, medians, and proportions. Based on this, I will impute missing values in medical_history_2 and medical_history_5 with the mode of each variable. 
This is because the subset of rows containing NA values seem to be a random sample of the entire subset i.e. missing completely at random, rather than missing not at random or missing at random. 

Limitations further discussed later.       

Impute missing data 
```{r}
# medical_history_2 - mode = 0 
# medical_history_5 - mode = 0 
final_df[is.na(final_df)] = as.factor(0)
summary(final_df)  
str(final_df) 
```

```{r}
boxplot(final_df$lab_result_1)
boxplot(final_df$lab_result_2)
boxplot(final_df$lab_result_3)
```


### Analysis of cost drivers ###  

1. Correlation
```{r}
df_cor = select(final_df, total_bill, total_num_admissions, lab_result_1, lab_result_2, lab_result_3, weight, height, len_stay, BMI, admission_age, nth_admission)  
```

```{r}
cor(df_cor)
```
The correlation analysis does not seem to produce any valuable insight. Attempt other methods instead. 

2. Feature importance  
```{r}
df_wo_id = select(final_df, -patient_id)
```

```{r}
library(randomForest)
library(caret)
library(varImp)
```

Random forest
```{r}
rf_model = randomForest(total_bill ~ ., data = df_wo_id, importance = TRUE)
```

```{r}
varImpPlot(rf_model) 
``` 
 
Top 3 variables: symptom 5, resident status, race  


### Visualisation ###
```{r}
library(ggplot2)
```

1. symptom_5 
```{r}
View(final_df %>% select(total_bill, symptom_5))
```

```{r}
tapply(final_df$total_bill, final_df$symptom_5, mean) 
```
```{r}
tapply(final_df$total_bill, final_df$symptom_5, median) 
```
Patients with symptom 5 have higher mean and median bills. 

```{r}
quantile(final_df$total_bill, prob = seq(0,1,0.1))
```

```{r}
# 0 to <10
table(final_df$symptom_5[final_df$total_bill < 11112.73])

# 10 to <20
table(final_df$symptom_5[final_df$total_bill >= 11112.73 & final_df$total_bill < 13577.63]) 

# 20 to <30
table(final_df$symptom_5[final_df$total_bill >= 13577.63 & final_df$total_bill < 15919.81]) 

# 30 to <40
table(final_df$symptom_5[final_df$total_bill >= 15919.81 & final_df$total_bill < 18047.39]) 

# 40 to <50
table(final_df$symptom_5[final_df$total_bill >= 18047.39 & final_df$total_bill < 20180.02]) 

# 50 to <60
table(final_df$symptom_5[final_df$total_bill >= 20180.02 & final_df$total_bill < 22598.71]) 

# 60 to <70
table(final_df$symptom_5[final_df$total_bill >= 22598.71 & final_df$total_bill < 25172.88])

# 70 to <80
table(final_df$symptom_5[final_df$total_bill >= 25172.88 & final_df$total_bill < 28384.32])

# 80 to <90
table(final_df$symptom_5[final_df$total_bill >= 34207.79 & final_df$total_bill < 98723.85]) 

```

```{r}
prop_s5 = c(4/340,
           31/340,
           82/340,
           141/340,
           195/340,
           221/340,
           248/340,
           268/340,
           297/340,
           303/339) 
quantile_bill = seq(0.1,1,0.1)
```

```{r}
ggplot(data = NULL, aes(x=quantile_bill, y=prop_s5)) +
  geom_point() +
  geom_line() +
  labs(title="Correlation between display of symptom 5 & total bill", x="Total bill (by quantile)", y="Proportion of patients with symptom 5") + 
  theme(plot.title = element_text(hjust = 0.5))
```
2. Resident status
```{r}
table(final_df$resident_status)
```

```{r}
tapply(final_df$total_bill, final_df$resident_status, mean)
```

```{r}
tapply(final_df$total_bill, final_df$resident_status, median) 
```


```{r}
ggplot(data = final_df, aes(x = resident_status, y = total_bill)) +
  geom_boxplot() +
  labs(x = "Resident Status", y = "Total Bill")
```
3. Race
```{r}
table(final_df$race)
```

```{r}
tapply(final_df$total_bill, final_df$race, mean)
```

```{r}
tapply(final_df$total_bill, final_df$race, median) 
```


```{r}
ggplot(data = final_df, aes(x = race, y = total_bill)) +
  geom_boxplot() +
  labs(x = "Race", y = "Total Bill")
```
Further analysis on Race - find other possible causes of discrepancies in bill for Malay & Indian races 
```{r}
ggplot(data = final_df, aes(x = race, y = len_stay)) +
  geom_boxplot() +
  labs(x = "Race", y = "Length of Stay")
```
```{r}
ggplot(data = final_df, aes(x = race, y = admission_age)) +
  geom_boxplot() +
  labs(x = "Race", y = "Age")
``` 
```{r}
rf_model_race = randomForest(race ~ ., data = df_wo_id, importance = TRUE)
```

```{r}
varImpPlot(rf_model_race) 
``` 

4. Symptom 5 to race 
```{r}
table(final_df$symptom_5, final_df$race) 
```
For each race, not significantly imbalanced between 0 and 1.

5. Race to resident status
```{r}
table(final_df$resident_status, final_df$race) 
```
Find percentage of Foreigners in each race 
```{r}
f = c(106,     14,    31,     10)
pr = c(338,     43,   109,     25)
sg = c(1722,    287,   567,    148)
proportion = f/(f+pr+sg)
proportion
```

Similar proportion of Foreigners across races.


### Analysis ###
1. Biological differences
Some races could have higher predisposition to other symptoms not captured in this data set.
Other conditions that are related to this particular condition, making treatment more complicated and costly.

2. Cultural differences
Cultural factors that pertain to healthcare services & needs, e.g. differences in health-seeking behaviours, accessing appropriate healthcare services & treatment plans, and disparities in health insurance coverage. 


### Recommendations ###
1. Resource & manpower allocation
Deploy appropriate healthcare professionals who can bridge cultural gaps and contextualise treatment plans to better fit patients’ needs.
Involves greater levels of personalisation, which can be automated through the use of AI technologies and EMR.
  
2. Collaboration with community resources and networks
Improve care continuity, especially for chronic conditions. 
Increase access to appropriate resources (e.g. differing needs across races/ conditions) and improve health-seeking behaviours
Reduce the effects of peripheral/ related symptoms and prevent them from becoming debilitative or severe.

3. Alternative cost-effective treatment plans
Empathise with differing financial circumstances based on resident status, medical insurance coverage, income levels, etc. 
Make cheaper alternatives available while explaining similarities and differences in alterative treatment plans, should patients want to opt in. 


### Limitations ###
1. Handling NA values
Further statistical tests can be performed to determine the missing data mechanism (MCAR/ MAR/ MNAR). 
Use of mode to impute values: 

medical_history_5: high proportion of entries (94%) are 0, hence it seems relatively reasonable to impute NAs as 0 
medical_history_2: this is more problematic because there is actually a higher proportion (29%) of 1s

Ultimately, domain knowledge + more information is required.  For example, it might be possible to infer or make better assumptions on missing data based on patients' medical history for related variables. 
 
2. Feature engineering
Length from last discharge: back to back admissions (e.g. within 90 days) could be an indication of something more serious, and might therefore drive costs up. 
However, for this data set, only 23% of entries are repeat admissions, hence it did not seem very useful. Furthermore, based on feature importance, the total number of admissions and repeated admissions did not play a significant importance in determining costs. 

3. Consider runner up variables in the feature importance plot
